Interactive Filter and Display of Hillary Clinton's Emails: \\A Cautionary Tale of Metadata

### 2017-04-10 VAST_submission/vgtc_journal_latex/paper.tex
Cleaning up text

On March 16, 2016 Wikileaks  published a searchable archive \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton on her private server.  The documents were provided as pdfs by the U.S. Department of State in response to Freedom of Information Act (FOIA) \cite{FOIA} requests.  The State Dept. also provided a searchable web archive of the documents, released in several instalments from May 2015 to March 2017\cite{StateDeptFOIA}.   Both sites provide a  useful tool for anyone searching for particular terms in the documents.  The Wikileaks site was put to much use by investigative reporters, and others, to search for topical news items.

What is critically missing from either site is a facility to learn summary, or statistical, features across all, or groups of,  emails.   A separate site where visual displays of summary features of the emails, which can be filtered by user selection, will go a long way towards filling this need.  The site would be \emph{complementary} to the more typical ``content search and record display'' site as provided by either online archive.   Add general internet search in a third browser, and the three combined would provide the means to actively explore the contents of any email archive.    One imagines, for example, having three browsers open simultaneously, one for the data base search (e.g. Wikileaks), one for statistical visualization and exploration (proposed here), and one for internet search (to provide important context for the other two services).  Each fills a need not met by the other two; the synergy of all three provide the user a powerful set of investigative tools.

In this paper, we describe an implementation of such a visual analytic service and illustrate it on the Secretary Clinton's e-mail archive.  The implementation is located at  \texttt{rshiny.math.uwaterloo.ca/clinton} where any visitor may interactively explore summary features  of the entire corpus of emails.   Different visual displays present different salient features from all selected emails; emails can be selected via a variety of interactive filters.   All displays are reactive and update simultaneously in response to user interaction with the data filters.  Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in the emails.

The paper is organized as follows.  We provide a brief timeline on the private email server to provide sufficient context for the work.  Readers more familiar with the background may wish to skip to the next three sections where the implementation is described.  Section \ref{sect:data} discusses the source of the data, its extraction and cleaning,  details of the metadata,  redaction codes as quasi-metadata,  and our limited use of the actual email content. 
Section \ref{sect:Displays} describes the displays used and illustrates them on the data. These include a spoked network plot, an email volume time series plot, a scatterplot of daily sending times, a barplot of exemption codes, and term frequency displays. Section \ref{sect:Filters} presents the interactive filters that can be applied to change the displays.  With the implementation described, some interactive analyses of the Clinton email archive are conducted to give a quick overview of the utility of the service.  We find some interesting, and potentially contentious, patterns in the email correspondence.  The section closes with  caveats on over interpretation of these analyses.  Some closing thoughts are given in the final section.

@@

On April 12, 2015 Clinton announces that she is running for the U.S. Presidency.  On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce that they found classified information in the emails and that the information was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact. On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.

Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton for her use of a private email server \cite{nochargeFBI}.  On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin had been found on his laptop.  Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.  In both cases, Comey and the FBI are criticized by pundits from different political parties.

\section{The data}
\label{sect:data}
As of March 3, 2017, a total of 32,795 available emails, either to or from Hillary Clinton, have been made publicly available in pdf form as a searchable database \cite{StateDeptFOIA}.  Many of these have been redacted according to the FOIA exemption codes \cite{FOIA}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
\caption{Inner Circle over all correspondence}
\label{fig:InnerCircleAll}
\end{center}
\end{figure}
Figure~\ref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.   The other red account is that of Sidney Blumenthal.  This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to provide Clinton advice throughout her tenure.  Wendy Sherman appears as orange. Sherman was appointed by Clinton to be Under Secretary of State for Political Affairs in 2011 and the high frequency of Sherman emails dates from this time.  Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.  Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided \cite{WikileaksPull}.

The width of each spoke is proportional to volume of emails between correspondents, the wider is their connecting spoke.  Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.  As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
\subsection{Email volume}
\label{sect:Displays:volume}
Figure~\ref{fig:VolumeAll} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{VolumeFullTime}
\caption{Email Volume: All emails (top) ; Those sent by Clinton (bottom)}
\label{fig:VolumeAll}
\end{center}
\end{figure}
presents two time series showing the volume of emails for each day.  In grey at top is the daily total; in magenta at bottom is the total for the subset chosen by the filters.  In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails sent from Clinton to others.

@@

\subsection{Email times}
\label{sect:Displays:times}
Figure~\ref{fig:TimesAll}  
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{DailyFullTimeFromClinton}
\caption{Email time patterns: all emails sent from Clinton}
\label{fig:TimesAll}
\end{center}
\end{figure}
shows all emails sent by Clinton over the four year period.  Each email appears as a point determined by the time shown that it was sent.  The calendar date determines the position on the horizontal axis,  the (24hr) time of day determines the vertical.   Points are coloured red if any part of that email was redacted, and blue if it was released in full.   Alpha blending is used to minimize the effect of over-plotting.

@@

If these are the actual times Secretary Clinton composed and sent email, it suggests that much of that activity occurred in the middle of the night.  To reflect the low email activity period as actual ``downtime'' between ``days'' (now interpreted as separable 24 hour email cycles),  we offer the possibility of choosing 6 PM as the ``end'' of a day rather than midnight.   This places the least active email period as a ``quiet'' boundary between days.   Unfortunately, though it better represents the rhythm of correspondence, it can be confusing since it means that the ``next day''  begins just after 6 PM.

Note also the surprisingly regular horizontal lines that appear across the top of the plot.  This regularity is suggestive of some automated schedule for the server which causes the email to stack up before being recorded at either 2 or 3 AM.  The location of these lines switches exactly whenever daylight savings time switches in North America.

@@

\subsection{FOIA exemptions}
\label{sect:Displays:FOIA}
U.S. Freedom of Information Act (FOIA) exemption codes B1 through B9 are used in each place in an email where information has been redacted.  For the authoritative definition the act itself should be consulted \cite{FOIA}.  Briefly, they mark each redaction as B1 for national security and foreign policy matters,  B2 for personnel practices, B3 for statutory exemptions,  B4 for trade secrets or financial information obtained in confidence,  B5 for inter- or intra-agency memorandums,  B6 for personal privacy,  B7 for records compiled for law enforcement,  B8 for records prepared in relation to financial monitoring institutions, and B9 for geophysical and geological information concerning oil and gas wells.   Each email can contain redactions for any number of exemption codes.

Figure \ref{fig:FOIAcodes}
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{ExemptionsFullTime}
\caption{FOIA barplot: Exemption codes from all email}
\label{fig:FOIAcodes}
\end{center}
\end{figure}
is a barplot of all FOIA exemption codes found in all of the emails.  As can be seen, B1 redactions made for national security and foreign policy reasons account for a small portion of all redactions made though affect more than 2,000 emails in the collection.   Nearly half of the emails contain B6 exemptions, those made to protect personal privacy.   The inter-agency, intra-agency exemption B5 is used in more than 8,000 emails.  This code has come under some criticism from advocates for transparent government as being overly applied \cite{unredactedB5}.  The remaining codes appear in very few emails and about 12,000 emails, or about a third of them, are not redacted at all.

As with all other displays, the FOIA barplot will update in reaction to all filters making it a useful tool to identify how codes co-occur over any selection of emails.  Conversely, filtering on FOIA codes can be interesting to see who sees correspondence of varying sensitivities, especially B1 and B5.

@@

The terms give some limited insight into the contents of the email.  As seen in Figure~\ref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.  One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies which could magnify the within email frequency for some terms.

As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is also challenged by this messy and non-standard data. The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'. The stemming algorithm was also challenged by typographical errors in emails and the proliferation of acronyms (e.g. for individuals and government abbreviations).

@@

This filter is simple but powerful.  All emails are displayed when the range covers all dates.  Moving the end points towards one another allows the user to focus the displays on any particular range, down to as fine as all emails on a single day.  With a fixed range of days, for example a two week period,  dragging the middle bar from left to right will have each display smoothly update over time.
%
%Time filtering within the app can be completed at the level of individual days. The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days. As well, clicking and dragging the blue bar in between the boundaries allows the user to sweep the data with an interval of a fixed width. Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
\subsection{Sent or received}
A drop down menu 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.35\linewidth]{ToFromImage}
\caption{Filtering on whether Clinton sent or received the email, or either}
\label{fig:toFromFilter}
\end{center}
\end{figure}
is used to select those emails which Clinton sent, or received, or either sent or received.  The default is ``All Emails'', which is the same as no filter being applied for sent or received.  Choosing ``From Clinton'' selects for display only those mails sent by Secretary Clinton, choosing ``To  Clinton'' selects only those emails which she received.  This might be used, for example, to explore whether the inner circle of correspondents changes depending on whether Secretary Clinton is emailing them, or they are emailing her.

\subsection{FOIA exemptions}
Multiple filtering by FOIA exemption codes is supported using a box selection area. 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.55\linewidth]{FOIABoxImage}
\caption{FOIA codes:  Emails containing any of these codes appear}
\label{fig:selectFOIA}
\end{center}
\end{figure}
Figure~\ref{fig:selectFOIA} displays the codes that are selected for email display.  Any email having one or more of these codes will appear in the displays.  To not have emails display with a B1 exemption code, for example, then the user selects and deletes the B1 from the list.  Or to have only the emails that  contain B1 redactions displayed, all other codes can deleted from the list.  Codes are returned to active display at any time by clicking in the display box and selecting them again from a drop down menu of the excluded codes.  Any number of codes  (one or more) can be selected to be included in the displays.

\subsection{Auxiliary information}
To demonstrate how other sources of information can easily be used to supplement the metadata found in the email archive, we consulted the State Department website and retrieved Secretary Clinton's official foreign travel schedule  \cite{ForeignSched}.  This information was added to the display as a simple check box ``Display Foreign Travel Schedule''.  When checked, light blue vertical bars appear on the  email volume time series display.  Appearing underneath the series, these mark time periods when Secretary Clinton was travelling outside the U.S. on State Department business.

\section{Some interactive analyses}
\label{sect:Analysis}
Each of the displays and filters described in the previous sections gives a slightly different and informative view of summary features of the email archive.  However, their analytic power is considerably amplified when used together.  Rather complex queries can be formed involving date, classification code, and whether Clinton is sending or receiving email.  Moreover, given a particular pattern in one display, its relation to that in another can be of considerable interest.   Together with access to the archive contents and an internet search engine, an enormous amount can be learned from these visual analytic tools.

In this section, we explore a few avenues of enquiry that might naturally arise.  At best this only hints at the analytic power that interactive filtering and display can bring to bear on an email archive.  To get a better appreciation, the user is encouraged to try it themselves at \texttt{rshiny.math.uwaterloo.ca/clinton}.  The tool is useful both for generating and for testing hypotheses.

\subsection{The last two years}
One might begin an interactive analysis by looking at the email patterns in the last two years of Clinton's tenure as Secretary of State.  
Figure~\ref{fig:lastTwoYears} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LastTwoYears}
\caption{Last two years:  Selected Emails are those that are B1 redacted}
\label{fig:lastTwoYears}
\end{center}
\end{figure}
shows the email volume time series over this period. 
This is accomplished by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.   Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.  The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.  As can be seen, there are relatively few of these.

Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).   The next largest spikes appear at the far left of this time window, at the beginning of 2011.  We might choose to compress the time window more, to focus on this section of the emails.

Bringing the right end of the time slider in to April 15, 2011 produces the series of Figure~\ref{fig:early2011B1}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1JanApril2011}
\caption{Jan. 1 -- Apr 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:early2011B1}
\end{center}
\end{figure}
As can be seen, there continue to be highs and lows of email activity.   The highest peak now is at the far left.  This is January 29, 2011, the day after President Mubarak of Egypt has ordered the army into the streets of Cairo to quell protests \cite{cairoTanks2011, cairoResponse2011}.  Dipping into the contents of the Wikileaks email archive, much of the email on this day is seen to be about managing the Egyptian Crisis.

Slightly to the right of this peak, there are several more.   At this time, international forces are gathering to consider Libya and its leader Muammar Gaddafi.  Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on the Libyan crisis, we move the sliders in to focus on the time range from February 25 when Secretary Clinton announced the suspension of the Libyan embassy in Washington, to March 18, the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes. 1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}.  On March 19, 2011 NATO military operations in Libya began \cite{LibyaTimelineWiki}.

The readjusted time series are shown in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LibyaBuildup}
\caption{Feb. 25 -- March 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1LibyaBuildup}
\end{center}
\end{figure}
The left most spike is the rightmost spike of approximately 80 emails from Figure~\ref{fig:early2011B1}.   
Of course all other displays have been updating every time the time slider is adjusted.

@@

Third, it is important that users realize that the value of metadata is not just in itself, where one might expect to easily discern general day-to-day habits such as one's circle of correspondents.  Rather it is that when coupled with other sources, which are abundantly and publicly available for Senator Clinton, much more can be learned than from any one source alone.  Those who have knowledge of, or access to, other sources may filter metadata to test previously held hypotheses; conversely, exploration of metadata could uncover patterns that generated hypotheses to be tested elsewhere.

>>>>

On March 16, 2016 Wikileaks  published a searchable archive \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton on her private server.  The documents were provided as pdfs by the U.S. Department of State in response to Freedom of Information Act (FOIA) \cite{FOIA} requests.  The State Dept. also provided a searchable web archive of the documents, released in several instalments from May 2015 to March 2017\cite{StateDeptFOIA}.   Both sites provide a  useful tool for anyone searching for particular terms in these documents.  The Wikileaks site was put to much use by investigative reporters, and others, to search for topical news items.

What is critically missing from either site is a facility to learn summary, or statistical, features across all, or groups of,  emails.   A separate site where visual displays of summary features of the emails, which can be filtered by user selection, would go a long way towards filling this need.  The site would be \emph{complementary} to the more typical ``content search and record display'' site as provided by either online archive.   Add general internet search in a third browser, and the three combined would provide the means to actively explore the contents of any email archive.    One imagines, for example, having three browsers open simultaneously, one for the data base search (e.g. Wikileaks), one for statistical visualization and exploration (proposed here), and one for internet search (to provide important context for the other two services).  Each fills a need not met by the other two; the synergy of all three provide the user a powerful set of investigative tools.

In this paper, we describe an implementation of such a visual analytic service and illustrate it on the Secretary Clinton's e-mail archive.  The implementation is located at  \texttt{rshiny.math.uwaterloo.ca/clinton} where any visitor may interactively explore summary features  of the entire corpus of emails.   Different visual displays present different salient features from all selected emails; emails can be selected via a variety of interactive filters.   All displays are reactive and update simultaneously in response to user interaction with the data filters.  Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in, the emails.

The paper is organized as follows.  We provide a brief timeline on the private email server to provide sufficient context for the work.  Readers more familiar with the background may wish to skip to the next three sections where the implementation is described.  Section \ref{sect:data} discusses the source of the data, its extraction and cleaning,  details of the metadata,  redaction codes as quasi-metadata,  and our limited use of the actual email content. 
Section \ref{sect:Displays} describes the displays used and illustrates them on the data. These include a spoked network plot, an email volume time series plot, a scatterplot of daily sending times, a barplot of exemption codes, and term frequency displays. Section \ref{sect:Filters} presents the interactive filters that can be applied to change the displays.  With the implementation described, some interactive analyses of the Clinton email archive are conducted to give a quick overview of the utility of the service.  We find some interesting, and potentially contentious, patterns in the email correspondence and check these against information found from Google searches for the dates in which the patterns occur.  The section closes with caveats on over interpretation of these analyses.  Some closing thoughts are given in the final section.

@@

On April 12, 2015 Clinton announced that she would be running for the U.S. Presidency.  On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announced that they had found classified information in the emails and that the information would have been considered classified at the time it was sent \cite{serverClassified}.  The Clinton presidential campaign declared that the emails must have been classified after the fact. On August 19, 2015, Clinton described the allegation of her mishandling classified information as simply a ``disagreement between agencies'' \cite{clintonDenialGuardian}.

Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton for her use of a private email server \cite{nochargeFBI}.  On October 28, 2016, Comey revealed that, in a separate investigation into former Congressman Anthony Weiner, emails belonging to his wife Huma Abedin had been found on his laptop.  Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.  In the first case Comey and the FBI were criticized by republican pundits and in the second case by democratic pundits.

\section{The data}
\label{sect:data}
As of March 3, 2017, a total of 32,795 available emails, either to or from Hillary Clinton, have been made publicly available in pdf form as a searchable database \cite{StateDeptFOIA}.  Many of these have been redacted according to the Freedom of Information Act (FOIA) exemption codes \cite{FOIA}.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
\caption{Inner Circle over all correspondence}
\label{fig:InnerCircleAll}
\end{center}
\end{figure}
Figure~\ref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.   The other red account is that of Sidney Blumenthal.  This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to have provided Secretary Clinton advice throughout her tenure.  Wendy Sherman appears as orange. Sherman was appointed by Clinton to be Under Secretary of State for Political Affairs in 2011 and the high frequency of Sherman emails dates from this time.  Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.  Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided by the WikiLeaks HTML \cite{WikileaksPull} (e.g. occasionally, a blank header appeared in the WikiLeaks HTML that was not present in the pdf; we did not correct for these manually).

The width of each spoke is proportional to the volume of emails between correspondents.  Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.  As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
\subsection{Email volume}
\label{sect:Displays:volume}
Figure~\ref{fig:VolumeAll} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{VolumeFullTime}
\caption{Email Volume: All emails (top) ; Those sent by Clinton (bottom)}
\label{fig:VolumeAll}
\end{center}
\end{figure}
presents two time series showing the volume of emails for each day.  In grey at top is the daily total; in magenta at bottom is the total for the subset selected by the filters.  In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails that were sent \emph{from} Clinton.

@@

\subsection{Email times}
\label{sect:Displays:times}
Figure~\ref{fig:TimesAll}  
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{DailyFullTimeFromClinton}
\caption{Email time patterns: all emails sent from Clinton}
\label{fig:TimesAll}
\end{center}
\end{figure}
shows all emails sent by Clinton over the four year period.  Each email appears as a point determined by the time that it was sent.  The calendar date determines the position on the horizontal axis,  the (24hr) time of day determines the vertical.   Points are coloured red if any part of that email was redacted, and blue if it was released in full.   Alpha blending is used to minimize the effect of over-plotting.

@@

If these are the actual times Secretary Clinton composed and sent email, it suggests that much of that activity occurred in the middle of the night.  We might alternatively choose the low email activity period to mark the actual ``downtime'' between ``days'' (now interpreted as low-activity separated 24 hour email cycles);   to this end, it is possible to select 6 PM as the ``end'' of a 24 hour email cycle rather than midnight.   Though it better represents the rhythm of correspondence, this definition of the daily email cycle is unnatural to most people and so could be confusing to some users (hence by default the break occurs at midnight).

Note also the surprisingly regular horizontal lines that appear across the top of the plot.  This regularity is suggestive of some automated schedule for the server which causes the email to stack up before being recorded at either 2 or 3 AM.  The switch between these two times matches exactly the dates where daylight savings time switches in North America.

@@

\subsection{FOIA exemptions}
\label{sect:Displays:FOIA}
U.S. Freedom of Information Act (FOIA) exemption codes B1 through B9 are used anywhere in an email where information has been redacted.  For the authoritative definition of the codes the act itself should be consulted \cite{FOIA}.  Briefly, a redaction is marked as B1 for national security and foreign policy matters,  B2 for personnel practices, B3 for statutory exemptions,  B4 for trade secrets or financial information obtained in confidence,  B5 for inter- or intra-agency memoranda,  B6 for personal privacy,  B7 for records compiled for law enforcement,  B8 for records prepared in relation to financial monitoring institutions, and B9 for geophysical and geological information concerning oil and gas wells.   Each email can contain redactions for any number of exemption codes.

Figure \ref{fig:FOIAcodes}
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{ExemptionsFullTime}
\caption{FOIA barplot: Exemption codes from all email}
\label{fig:FOIAcodes}
\end{center}
\end{figure}
is a barplot of all FOIA exemption codes found in all of the emails.  As can be seen, B1 redactions made for national security and foreign policy reasons account for a small portion of all redactions made,  though this amounts to more than 2,000 emails in the collection.   Nearly half of the emails contain B6 exemptions, those made to protect personal privacy.   The inter-agency, intra-agency exemption B5 is used in more than 8,000 emails.  This code has come under some criticism from advocates for transparent government as being overly applied \cite{unredactedB5}.  The remaining codes appear in very few emails and about 12,000 emails, or about a third of them, are not redacted at all.

As with all other displays, the FOIA barplot will update in reaction to all filters and so provides a useful tool to identify co-occurrence of codes over any selection of emails.  Conversely, filtering on FOIA codes can be interesting to see who is privy to correspondence of varying sensitivities, especially B1 and B5.

@@

The terms give some limited insight into the contents of the email.  As seen in Figure~\ref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.  One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies; this could magnify the within email frequency for some terms.

As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is challenged by this messy and non-standard data. The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'. The stemming algorithm was also challenged by typographical errors in emails and the proliferation of acronyms (e.g. for individuals and government abbreviations).

@@

This filter is simple but powerful.  All emails are displayed when the range covers all dates.  Moving the end points towards one another allows the user to focus the displays on any particular range, down to as fine as all emails on a single day.  With a fixed range of days, for example a two week period,  dragging the middle bar from left to right will have each display smoothly update over time.
%
%Time filtering within the app can be completed at the level of individual days. The use of the mouse to click and drag the slider boundaries allows for large, rapid changes in the date range specified, while clicking on one of the boundaries and using the arrow keys allows for fine selection of the date range via single days. As well, clicking and dragging the blue bar in between the boundaries allows the user to sweep the data with an interval of a fixed width. Clicking the middle bar and using the arrow keys allows the user to shift the range selected by individual days, analogously to the boundaries.
\subsection{Sent or received}
A drop down menu 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.35\linewidth]{ToFromImage}
\caption{Filtering on whether Clinton sent or received the email, or either}
\label{fig:toFromFilter}
\end{center}
\end{figure}
is used to select those emails which Clinton sent, or received, or either (i.e. sent or received).  The default is ``All Emails'', which is the same as no filter being applied for sent or received.  Choosing ``From Clinton'' selects for display only those mails sent by Secretary Clinton, choosing ``To  Clinton'' selects only those emails which she received.  This might be used, for example, to explore whether the inner circle of correspondents changes depending on whether Secretary Clinton is emailing them, or they are emailing her.

Note that if neither the ``From:''  nor ``To:''  were captured from the HTML, then such emails can appear only when ``All Emails'' is selected.

\subsection{FOIA exemptions}
Multiple filtering by FOIA exemption codes is supported using a box selection area. 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.55\linewidth]{FOIABoxImage}
\caption{FOIA codes:  Emails containing any of these codes appear}
\label{fig:selectFOIA}
\end{center}
\end{figure}
Figure~\ref{fig:selectFOIA} displays the codes that are selected for email display.  Any email having one or more of these codes will appear in the displays.  To not have emails display with a B1 exemption code, for example, then the user selects and deletes the B1 from the list (multiple selection is enabled).  Or to have only the emails which  contain B1 redactions displayed, all other codes can deleted from the list.

Codes are returned to active display at any time by clicking in the display box and selecting them again from a drop down menu of the excluded codes.  Any number of codes  (one or more) can be selected to be included in the displays.

\subsection{Auxiliary information}
To demonstrate how other sources of information can easily be used to supplement the metadata found in the email archive, we consulted the State Department website and retrieved Secretary Clinton's official foreign travel schedule  \cite{ForeignSched}.  This information was added to the display as a simple check box ``Display Foreign Travel Schedule''.  When checked, light blue vertical bars appear on the  email volume time series display.  Appearing behind the series, these mark time periods when Secretary Clinton was travelling outside the U.S. on State Department business according to the published schedule.

\section{Some interactive analyses}
\label{sect:Analysis}
Each of the displays and filters described in the previous sections gives a slightly different informative view of summary features of the email archive.  Their analytic power is considerably amplified when used together.

Rather complex queries can be formed involving date, classification code, and whether Clinton is sending or receiving email.  Moreover, given a particular pattern in one display, its relation to that in another can be of considerable interest.   Together with access to the archive contents and an internet search engine, an enormous amount can be learned from these visual analytic tools.

In this section, we explore a few avenues of enquiry that might naturally arise.  At best these only hint at the analytic power that interactive filtering and display can bring to bear on an email archive.  The user is encouraged to try it themselves at \texttt{rshiny.math.uwaterloo.ca/clinton}to get a fuller appreciation of the power of the tool.  Those who are exploring the corpus will find it useful for generating hypotheses that might be tested against other sources.   This is the approach taken here.  Conversely, those more knowledgeable about the contexts surrounding Secretary Clinton and her private email server might use the visual analytic tool more as a means to test hypotheses already held.

\subsection{The last two years}
We begin our interactive analysis by looking at  email patterns in the last two years of Clinton's tenure as Secretary of State.

Figure~\ref{fig:lastTwoYears} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LastTwoYears}
\caption{Last two years:  Selected Emails are those that are B1 redacted}
\label{fig:lastTwoYears}
\end{center}
\end{figure}
shows the email volume time series over this period. 
This is accomplished by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.   Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.  The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.  As can be seen, there are \emph{relatively} few of these.

Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).   The next largest spikes appear at the far left of this time window, at the beginning of 2011.  We might choose to compress the time window more, to focus on the early 2011 section of the emails.  Bringing the right end of the time slider in to April 15, 2011 produces the series of Figure~\ref{fig:early2011B1}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1JanApril2011}
\caption{Jan. 1 -- Apr 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:early2011B1}
\end{center}
\end{figure}

As can be seen, there continue to be highs and lows of email activity.   The highest peak now is at the left, though closer to the centre of Figure~\ref{fig:early2011B1} than it had been in Figure~\ref{fig:lastTwoYears} .  This is January 29, 2011, the day after President Mubarak of Egypt has ordered the army into the streets of Cairo to quell protests \cite{cairoTanks2011, cairoResponse2011}.  Dipping into the contents of the Wikileaks email archive, much of the email on this day is seen to be about managing this Egyptian crisis.

Just to the right of this peak, there are several more.   At this time, international forces are gathering to consider Libya and its leader Muammar Gaddafi.  Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on the Libyan crisis, we move the sliders in again, this time to focus on the time range from February 25 (when Secretary Clinton announced the suspension of the Libyan embassy in Washington) to March 18 (the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes. 1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}).  On March 19, 2011 NATO military operations in Libya began \cite{LibyaTimelineWiki}.

The readjusted time series are shown in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LibyaBuildup}
\caption{Feb. 25 -- March 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1LibyaBuildup}
\end{center}
\end{figure}
The left most spike in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}  is the rightmost spike of the pair of approximately 80 emails from Figure~\ref{fig:early2011B1}.   
Of course all other displays have been updating every time the time slider is adjusted.

@@

Third, it is important that users realize that the value of metadata is not just in itself, where one might expect to easily discern general day-to-day habits such as one's circle of correspondents.  Rather it is that when coupled with other sources, which are abundantly and publicly available for Secretary Clinton, much more can be learned than from any one source alone.  Those who have knowledge of, or access to, other sources may filter metadata to test previously held hypotheses; conversely, exploration of metadata could uncover patterns that generated hypotheses to be tested elsewhere.
